### 2022-01-28 .debris/2022-01-27 11.00.19.00/Paper.tex
modified

\documentclass[hidelinks,10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[cm]{fullpage}
\usepackage[round]{natbib}
\bibliographystyle{abbrvnat}
\usepackage{doi}
\usepackage{bm}
\usepackage[table]{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{float}
\restylefloat{table}
\usepackage{siunitx}
\DeclareSIUnit\year{yr}

@@

\section{Introduction}\label{sec:intro}
This document includes a description of the code FALCON, with a presentation of the features implemented followed by results of extensive benchmarks listed in
Table \ref{tab:list}, performed to verify that the code solves correctly a wide range of problems. Description and results of all benchmarks are shown
in Section \ref{sec:benchmark}.

\begin{table}[H]
\caption{List of benchmarks performed for each feature of the code. Data of all benchmarks can be found at 
\url{https://github.com/aleregorda/Benchmarks}}
\centering
\begin{tabular}{| l | l | l |}
\toprule
 \textbf{Feature}  & \textbf{Benchmark} & \textbf{Section} \\
\midrule
  Solver  & Stokes flow & \ref{sec:stokes} \\
\hline
  \multirow{2}{*}{Markers advection} & Zalesak disk & \ref{sec:runge} \\
    & Conservative Interpolation Velocity & \ref{sec:CVI} \\
\hline
  \multirow{4}{*}{Momentum equation}  & Poiseuille flow & \ref{sec:poiseuille}\\
    & Instantaneous 2D sphere & \ref{sec:ist_sphere} \\
    & Rayleigh-Taylor experiment & \ref{sec:rayleigh} \\
    & Falling block & \ref{sec:block} \\
\hline
    & 2D Stokes sphere & \ref{sec:sphere} \\
    Sticky air and  & Stabilisation algorithm & \ref{sec:stab} \\
    free surface & Topography relaxation & \ref{sec:crameri} \\
  & Spontaneous subduction & \ref{sec:subduction} \\
\hline
  Erosion and sedimentation  & - & - \\
\hline
   Non-linear  & Slab detachment & \ref{sec:slab} \\
   visco-plasticity & Indenter & \ref{sec:indenter} \\
    & Brick & \ref{sec:brick} \\
\hline
  \multirow{3}{*}{Energy equation}  & Advection stabilisation & \ref{sec:advection} \\
    & Simple shear heating & \ref{sec:simple_shear} \\
    & Shear and adiabatic heating & \ref{sec:shear} \\
\hline
\multirow{3}{*}{Energy + momentum} & Mantle convection & \ref{sec:mantle} \\
& Viscoplastic mantle convection & \ref{sec:visco_mantle} \\
& Thin layer entrainment & \ref{sec:thin} \\
\hline
  Phase changes  & \multirow{2}{*}{Hydrated sinking cylinder} & \multirow{2}{*}{\ref{sec:quinquis}} \\
  and hydration  &  & \\
\hline
  Melting & Experimental melting curves & \ref{sec:katz} \\
\bottomrule
\end{tabular}
\label{tab:list}
\end{table}

@@

The time step $dt$ is chosen in according to Courant-Friedrichs-Lewy condition \citep{Anderson1995}, such that
\[dt=C_n\; \min\left(\frac{\Delta x}{\max|\bm{u}|},\frac{(\Delta x)^2}{\kappa}\right)\]
where $0<C_n<1$ is the Courant (CFL) number, $\Delta x$ is the minimum dimension of smallest element, $\max|\bm{u}|$ is the maximum velocity calculated on the
entire domain and
$\kappa=\frac{k}{\rho C_p}$ is the thermal diffusivity \citep{Thieulot2014}.

@@

\section{Sticky air and free surface}\label{sec:sticky}
The Earth's surface can be treated by means of either the so-called sticky-air or a true free surface method, both of them implemented in the code. In the
sticky-air method the surface is approximated with the introduction of a buoyant layer with a viscosity at least four orders of magnitude lower than the crust
\citep{Schmeling2008,Crameri2012} and the interface between lithosphere and air is defined using a chain of passive markers that are advected as the Lagrangian
markers. The correctness of the evolution of the markers chain is tested with the experiment of a 2D time-dependent Stokes sphere below a free surface and
compared with results from ASPECT\footnote{\url{https://aspect.geodynamics.org/}} \citep{Kronbichler2012,Heister2017,Bangerth2020,Bangerth2020a} (Section
\ref{sec:sphere}). In the true free surface case the top boundary is assumed stress-free and velocities are not fixed. In this case topography variations are
described by vertical deformations of the mesh that depend on the velocity field of the nodes that identify the free surface, while horizontal deformations are
not taken into account. This procedure is known as the Arbitrary Lagrangian-Eulerian (ALE) method and its implementation follows the technique described in
\citet{Thieulot2011}. However, although the implementation of a true free surface better reproduces laboratory experiment, extremely small time steps can be
necessary to maintain stability \citep{Kaus2010a,Quinquis2011,Thieulot2014}. Therefore, the stability algorithm proposed by \citet{Kaus2010a} is implemented
to avoid instabilities due to high density differences at the free surface when using too large time steps. The implementation of this algorithm are tested by
performing the experiment described by \citet{Kaus2010a} (Section \ref{sec:stab}).

@@

\section{Erosion and sedimentation}\label{sec:erosion}
Surface processes at either the lithosphere-air interface or in correspondence of the free surface have been implemented by means of the software
Fastscape\footnote{\url{https://fastscape.org/}} \citep{Braun2013,Cordonnier2019,Yuan2019a,Yuan2019b}. FastScapeLib is a set of routines that solve the stream
power law (enriched by a sediment transport/deposition term) \citep{Yuan2019a}, hillslope diffusion and marine transport and deposition \citep{Yuan2019b}.
The partial differential equation solved by FastScapeLib is:
\begin{equation}
\frac{\partial h}{ \partial t}=U-K_f A^{m_e}S^{n_e} + \frac{G}{A}\int_A \left( U-\frac{\partial h}{\partial t} \right) dA + K_d \bm{\nabla}^2 h 
\end{equation}
where $h$ is the topography, $U$ is the uplift, $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $S$ is the slope, $m_e$ and $n_e$ are
parameters that should be chosen to constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}, $G$ is the dimensionless
deposition/transport coefficient for the enriched stream power law and $K_d$ is the effective diffusivity. Terms $K_f A^{m_e}S^{n_e}$ and $K_d \bm{\nabla}^2 h$
consider long-range processes, described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020}, and short-range processes,
proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}.

In case of sticky air, slope is determined by means of the markers chain, which is vertically corrected considering erosion and sedimentation rates. After the
correction, continental markers above the markers chain are transformed in air markers and air markers below the markers chain become sediments. Similarly,
slope is determined considering the nodes on the top boundary in case of true free surface and they are vertically corrected as for the markers chain. In this
case, continental markers above the top boundary are simply deleted.
%Surface processes at the lithosphere-air interface are implemented taking into account fluvial erosion (long-range processes), hillslope processes (short-range
%processes) and failure of slopes exceeding a threshold angle $\theta$ \citep{Burov1997,Tucker1999,Sacek2012,Croissant2014,Ueda2015,Beucher2020}. Erosion rates
%$\dot{E}_{lr}$ due to long-range processes can be described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020}, in according to
%\begin{equation}
%\dot{E}_{lr}=K_f A^{m_e} \bm{\nabla} h^{n_e}
%\end{equation}
%where $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $\bm{\nabla} h$ is the slope and $m_e$ and $n_e$ are parameters chosen in to
%constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}. Erosion rates $\dot{E}_{sr}$ due to short-range processes are
%implemented by assuming that transport is linearly proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}, in according to
%\begin{equation}
%\dot{E}_{sr}=K_d \bm{\nabla}^2 h
%\end{equation}
%where $K_d$ is the effective diffusivity. In addition, the slope is modified if it still exceeds a threshold angle after long- and short-range processes
%\citep{Ueda2015}.

@@

\section{Non-linear rheologies}\label{sec:nonlinear}
Non-linear rheologies are implemented combining viscous creep (dislocation and diffusion) and plastic yielding. For each marker, diffusion and dislocation
viscosity ($\eta_{df|ds}$) can be determined as follows \citep{Karato1993,Wang2016}:
\begin{equation}\label{eq:viscous}
\eta_{df|ds}=f_s \left(\frac{d^m}{A}\right)^{\frac{1}{n}} \left (\frac{2^{\frac{1-n}{n}}}{3^{\frac{n+1}{2n}}} \right)
I_2^{\frac{(1-n)}{n}}\exp\left(\frac{Q+pV}{nRT}\right)
\end{equation}
where $f_s$ is a scaling factor used to represent lithologies that are stronger or weaker than the base set, $d$ is the grain size, $m$ is the grain size exponent,
$n$ is the stress exponent, $A$ is the uniaxial pre-exponential factor, $I_2$ is the square root of the second invariant of the strain rate tensor, $Q$ is the
activation energy, $p$ is the pressure, $V$ is the activation volume, $R$ is the gas constant and $T$ is the temperature.
Pressures and temperatures are determined by the interpolation of the nodal parameters, while strain rates are calculated by means of the
derivative of the velocities on the nodes. In case of diffusion creep $n=1$ and $m>0$, in case of dislocation creep $n>1$ and $m=0$. The viscous creep $\eta_{cp}$
is then calculated as the harmonic average between $\eta_{df}$ and $\eta_{ds}$:
\begin{equation}\label{eq:visco_creep}
\eta_{cp}=\left(\frac{1}{\eta_{df}}+\frac{1}{\eta_{ds}}\right)^{-1}
\end{equation}
The implementation of the non-linear viscous creep viscosity depending on the strain rate is tested by means of the slab detachment benchmark
(Section \ref{sec:slab}).

@@

\section{Mantle melting}\label{sec:melt}
Melt fraction in the mantle depends on temperature ($T$, in \textdegree C), pressure ($p$, in \si{\giga\pascal}) and the water content in the melt ($X_{H_2O}$, in wt.\%).
The determination of the melt fraction $M$ is implemented as explained by \citet{Katz2003}, according to the successive modification by \citet{Langmuir2006} and
\citet{Kelley2010}, as follows
\begin{eqnarray}\label{eq:melt_fr}
M(p,T,X_{H_2O})&=&-T+T_{sol}+(x_t\cdot ln(p+y_t))\cdot X_{H_2O}-\nonumber\\
&&+K_m\left(\frac{X_{H_2O}^{bulk}}{D_{H_2O}(1-X_{H_2O})+X_{H_2O}}\right)^{\gamma}
\end{eqnarray}
where $T_{sol}$ is the temperature of the dry solidus, $D_{H_2O}$ is the partition coefficient and $x_t\cdot log(p+y_t)$ indicates the pressure dependence of
the melting curve \citep{Kelley2010}. $x_t$, $y_t$, $K_m$ and $\gamma$ are parameters chosen in according to \citet{Kelley2010}.
However, $X_{H_2O}$ depends on the melt fraction as
\begin{equation}\label{eq:melt_wt}
X_{H_2O}(M)=\frac{X_{H_2O}^{bulk}}{D_{H_2O}+M(1-D_{H_2O})}
\end{equation}
and a numerical solution can be found using a root-finder method \citep{Katz2003,Wang2016}. Furthermore, the content of water in the melt is limited by the
pressure-dependent saturation concentration of water in the melt, determined as
\begin{equation}\label{eq:melt_max}
X_{H_2O}^{sat}=\chi_1 p^{\lambda_m}+\chi_2 p
\end{equation}
where $\chi_1$, $\chi_2$ and $\lambda_m$ are parameters chosen according to \citet{Katz2003}, \citet{Langmuir2006} and \citet{Kelley2010}. The correctness in
the determination of melt fractions and water contents in the melt are tested comparing model results with experimental curves obtained by \citet{Katz2003}
(Sec. \ref{sec:katz}).

@@

\subsection{Instantaneous 2D sphere}\label{sec:ist_sphere}
The domain is a unit square with gravity $\bm{g}=(0,-1)$. The fluid has constant density and viscosity ($\rho_f=1$ and $\eta_f=1$). The sphere is in the middle
of the domain with a radius R = 0.123456798 and has constant density and viscosity ($\rho_s=10^{-2}+\rho_f$ and $\eta_s=10^3 \cdot \eta_f$).
We distinguish three different types of velocity boundary conditions:
\begin{enumerate}
\item FS: free slip conditions on all sides.
\item NS: no slip conditions on all sides.
\item OT: free slip conditions on the sides and the bottom, and open boundary at the top.
\end{enumerate}
All three cases are performed for different grid resolutions, between \num{16x16} and \num{512x512} elements with 50 randomly distributed markers per element,
and different average schemes for the viscosity (harmonic, geometric and arithmetic). The velocity in the centre of the sphere $\bm{v}(0.5;0.5)$), minimum and
maximum velocities ($u_{min}$, $u_{max}$, $v_{min}$, $v_{max}$) and pressures ($p_{min}$, $p_{max}$) on the entire domain, $v_{\textrm{rms}}$ and average pressure on the
entire domain ($p_{avg}$) are compared with solutions generated by ASPECT \citep{Kronbichler2012,Heister2017,Bangerth2020,Bangerth2020a} and all results
can be found at \url{https://github.com/cedrict/fieldstone/tree/master/images/stokes_sphere2D} and 
\url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum_equation/Instantaneous_2D_Sphere}. Results in terms of the $v_{\textrm{rms}}$ for NS model 
are also shown in Fig. \ref{fig:inst_sphere}.

@@

\subsection{2D time-dependent Stokes sphere with deformable free surface}\label{sec:sphere}
This experiment is performed in a unit square domain with the gravity acceleration fixed to $g_y=-1$. The sphere has $\rho_s=2$ and $\eta_s=10^3$ and is
initially centred at ($0.5;0.6$) with radius $R=0.123456789$. The fluid surrounding the sphere has $\rho_f=1$ and $\eta_f=1$ and occupies the domain for
$y\leq 0.75$, while for $y>0.75$ the air has $\rho_a=0$ and $\eta_a=10^{-3}$. Velocity boundary conditions are set to free slip on all sides. The Courant
number is set to 0.25. The experiments run for 200 s using grid resolutions from \num{150x150} to \num{512x512} elements, with 25 randomly distributed markers
per element, and for different average schemes for the viscosity. The interface between the fluid and the air is tracked by means of the markers chain. Results
in terms of velocity and pressure fields and topography variations are compared with results obtained with ASPECT
\citep{Kronbichler2012,Heister2017,Bangerth2020,Bangerth2020a} and they can be found at
\url{https://github.com/cedrict/fieldstone/tree/master/images/stokes_sphere_fs2D} and
\url{https://github.com/aleregorda/Benchmarks/tree/main/Surface_processes/Time_dependent_sphere}. Results in terms of the $v_{\textrm{rms}}$ in case of 
an arithmetic average are also shown in Fig. \ref{fig:inst_spherefs}.

@@

\subsection{Viscoplastic mantle convection}\label{sec:visco_mantle}
This benchmark is performed as Cases 1-5a described in \citet{Tosi2015} in a 2D unit square domain with two different grid resolutions of \num{32x32} 
and \num{100x100} elements. Reference viscosity, reference density, heat capacity, thermal conductivity and thermal expansion coefficient are fixed to a constant
value of 1 ($\eta_0= \rho_0= C_p = k=\alpha=1$), while the gravity has been chosen as $g_y=10^2$, in order to obtain a Rayleigh number $Ra=10^2$. Temperatures
are set to 0 on top and 1 on bottom of the domain. Velocity boundary conditions are set to free slip on all sides. The initial temperature field is given by
\[T(x,y)=(1-y)+0.01 \cos (\pi x) \sin (\pi y)\].
The viscosity field $\eta$ is calculated as the harmonic average between a linear part $\eta_{lin}$ that depends on temperature only or on temperature and 
depth $y$ and a nonlinear-plastic part $\eta_{pl}$ that depends on the strain rate $\dot{\bm{\epsilon}}$, as follows
\[\eta(T,y,\dot{\bm{\epsilon}})=2\left(\frac{1}{\eta_{lin}(T,y)}+\frac{1}{\eta_{pl}(\dot{\bm{\epsilon}})}\right)^{-1}\]
The linear and the nonlinear-plastic parts of the viscosity are calculated following \citet{Tosi2015} as follows
\[\eta_{lin}(T,y)=\exp(-\ln(\Delta\eta_T)T+\ln(\Delta\eta_y)y)\]
\[\eta_{pl}(\dot{\bm{\epsilon}})=\eta^*+\frac{\sigma_Y}{\sqrt{\dot{\bm{\epsilon}}:\dot{\bm{\epsilon}}}}\]
where $\Delta\eta_T$, $\Delta\eta_y$, $\eta^*$ and $\sigma_Y$ are parameters chosen as listed in Table \ref{tab:tosi_setup}.

@@

The experiment is performed with two grid resolution (\num{125x40} and \num{200x80} elements), with 100 markers per element and CFL of 0.25. $v_{\textrm{rms}}$
(see Sec. \ref{sec:error}) calculated for both the simulations match well with results from \citet{vanKeken1997} and obtained with ASPECT
\citep{Kronbichler2012,Heister2017,Bangerth2020,Bangerth2020a} and ELEFANT \citep{Thieulot2014}, obtained with similar grid resolutions
(Fig. \ref{fig:thin}). In addition, a variety of simulations with different resolutions and aspect ratios are performed to verify that the sensitivity of the
initial velocity field. Results match well with results from \citet{vanKeken1997} (in the grey area) and \citet{Thieulot2014} (Fig. \ref{fig:thin_initial}). 
All data can be found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum%2BEnergy/Thin_layer}.

>>>>

\documentclass[hidelinks,10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[cm]{fullpage}
\usepackage[round]{natbib}
\bibliographystyle{abbrvnat}
\usepackage{doi}
\usepackage{bm}
\usepackage[table]{xcolor}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}
\usepackage{float}
\restylefloat{table}
\usepackage{siunitx}
\DeclareSIUnit\year{yr}

@@

\section{Introduction}\label{sec:intro}
This supplement includes a description of the code, with a presentation of the features implemented followed by results of extensive benchmarks listed in
Table \ref{tab:list}, implemented to verify that the code solves correctly a large range of problems. Description and results of all benchmarks are shown
in Section \ref{sec:benchmark}.

\begin{table}[H]
\caption{List of benchmarks implemented for each feature of the code. Data of all benchmarks can be found at 
\url{https://github.com/aleregorda/Benchmarks}}
\centering
\begin{tabular}{| l | l | l |}
\toprule
 \textbf{Feature}  & \textbf{Benchmark} & \textbf{Section} \\
\midrule
  Solver  & Stokes flow & \ref{sec:stokes} \\
\hline
  \multirow{2}{*}{Markers advection} & Zalesak disk & \ref{sec:runge} \\
    & Conservative Interpolation Velocity & \ref{sec:CVI} \\
\hline
  \multirow{4}{*}{Momentum equation}  & Poiseuille flow & \ref{sec:poiseuille}\\
    & Instantaneous 2D sphere & \ref{sec:ist_sphere} \\
    & Rayleigh-Taylor experiment & \ref{sec:rayleigh} \\
    & Falling block & \ref{sec:block} \\
\hline
    & 2D Stokes sphere & \ref{sec:sphere} \\
    Sticky air and  & Stabilisation algorithm & \ref{sec:stab} \\
    free surface & Topography relaxation & \ref{sec:crameri} \\
  & Spontaneous subduction & \ref{sec:subduction} \\
\hline
  Erosion and sedimentation  & - & - \\
\hline
   Non-linear  & Slab detachment & \ref{sec:slab} \\
   visco-plasticity & Indenter & \ref{sec:indenter} \\
    & Brick & \ref{sec:brick} \\
\hline
  \multirow{3}{*}{Energy equation}  & Advection stabilisation & \ref{sec:advection} \\
    & Simple shear heating & \ref{sec:simple_shear} \\
    & Shear and adiabatic heating & \ref{sec:shear} \\
\hline
\multirow{3}{*}{Energy + momentum} & Mantle convection & \ref{sec:mantle} \\
& Viscoplastic mantle convection & \ref{sec:visco_mantle} \\
& Thin layer entrainment & \ref{sec:thin} \\
\hline
  Phase changes  & \multirow{2}{*}{Hydrated sinking cylinder} & \multirow{2}{*}{\ref{sec:quinquis}} \\
  and hydration  &  & \\
\hline
  Melting & Experimental melting curves & \ref{sec:katz} \\
\bottomrule
\end{tabular}
\label{tab:list}
\end{table}

@@

The time step $dt$ is chosen in according to Courant-Friedrichs-Lewy condition \citep{Anderson}, such that
\[dt=C_n\; \min\left(\frac{\Delta x}{\max|\bm{u}|},\frac{(\Delta x)^2}{\kappa}\right)\]
where $0<C_n<1$ is the Courant (CFL) number, $\Delta x$ is the minimum dimension of smallest element, $\max|\bm{u}|$ is the maximum velocity calculated on the
entire domain and
$\kappa=\frac{k}{\rho C_p}$ is the thermal diffusivity \citep{Thieulot2014}.

@@

\section{Sticky air and free surface}\label{sec:sticky}
The Earth's surface can be treated by means of either the so-called sticky-air or a true free surface method, both of them implemented in the code. In the
sticky-air method the surface is approximated with the introduction of a buoyant layer with a viscosity at least four orders of magnitude lower than the crust
\citep{Schmeling2008,Crameri2012} and the interface between lithosphere and air is defined using a chain of passive markers that are advected as the Lagrangian
markers. The correctness of the evolution of the markers chain is tested with the experiment of a 2D time-dependent Stokes sphere below a free surface and
compared with results from ASPECT\footnote{\url{https://aspect.geodynamics.org/}} \citep{KHB12,heister_aspect_methods2,aspect-doi-v2.2.0,aspectmanual} (Section
\ref{sec:sphere}). In the true free surface case the top boundary is assumed stress-free and velocities are not fixed. In this case topography variations are
described by vertical deformations of the mesh that depend on the velocity field of the nodes that identify the free surface, while horizontal deformations are
not taken into account. This procedure is known as the Arbitrary Lagrangian-Eulerian (ALE) method and its implementation follows the technique described in
\citet{Thieulot2011}. However, although the implementation of a true free surface better reproduces laboratory experiment, extremely small time steps can be
necessary to maintain stability \citep{Kaus2010a,Quinquis2011,Thieulot2014}. Therefore, the stability algorithm proposed by \citet{Kaus2010a} is implemented
to avoid instabilities due to high density differences at the free surface when using too large time steps. The implementation of this algorithm are tested by
performing the experiment described by \citet{Kaus2010a} (Section \ref{sec:stab}).

@@

\section{Erosion and sedimentation}\label{sec:erosion}
%Surface processes at the lithosphere-air interface are implemented taking into account fluvial erosion (long-range processes), hillslope processes (short-range
%processes) and failure of slopes exceeding a threshold angle $\theta$ \citep{Burov1997,Tucker1999,Sacek2012,Croissant2014,Ueda2015,Beucher2020}. Erosion rates
%$\dot{E}_{lr}$ due to long-range processes can be described by a stream power law equation \citep{Tucker1999,Ueda2015,Beucher2020}, in according to
%\begin{equation}
%\dot{E}_{lr}=K_f A^{m_e} \bm{\nabla} h^{n_e}
%\end{equation}
%where $K_f$ is the erodibility coefficient, $A$ is the upstream drainage area, $\bm{\nabla} h$ is the slope and $m_e$ and $n_e$ are parameters chosen in to
%constrain the ratio $m_e/n_e$ between 0.35 and 0.8 \citep{Croissant2014,Ueda2015,Beucher2020}. Erosion rates $\dot{E}_{sr}$ due to short-range processes are
%implemented by assuming that transport is linearly proportional to slope \citep{Burov1997,Ueda2015,Beucher2020}, in according to
%\begin{equation}
%\dot{E}_{sr}=K_d \bm{\nabla}^2 h
%\end{equation}
%where $K_d$ is the effective diffusivity. In addition, the slope is modified if it still exceeds a threshold angle after long- and short-range processes
%\citep{Ueda2015}.

@@

\section{Non-linear rheologies}\label{sec:nonlinear}
Non-linear rheologies are implemented combining viscous creep (dislocation and diffusion) and plastic yielding. For each marker, diffusion and dislocation
viscosity ($\eta_{df|ds}$) can be determined as follows \citep{Karato1993,Wang2016}:
\begin{equation}\label{eq:viscous}
\eta_{df|ds}=\left(\frac{d^m}{A}\right)^{\frac{1}{n}}I_2^{\frac{(1-n)}{n}}\exp\left(\frac{Q+pV}{nRT}\right)
\end{equation}
where $d$ is the grain size, $m$ is the grain size exponent, $n$ is the stress exponent, $A$ is the pre-exponential factor, $I_2$ is the square root of the
second invariant of the strain rate tensor, $Q$ is the activation energy, $p$ is the pressure, $V$ is the activation volume, $R$ is the gas constant and $T$
is the temperature. Pressures and temperatures are determined by the interpolation of the nodal parameters, while strain rates are calculated by means of the
derivative of the velocities on the nodes. In case of diffusion creep $n=1$ and $m>0$, in case of dislocation creep $n>1$ and $m=0$. The viscous creep $\eta_{cp}$
is then calculated as the harmonic average between $\eta_{df}$ and $\eta_{ds}$:
\begin{equation}\label{eq:visco_creep}
\eta_{cp}=\left(\frac{1}{\eta_{df}}+\frac{1}{\eta_{ds}}\right)^{-1}
\end{equation}
The implementation of the non-linear viscous creep viscosity depending on the strain rate is tested by means of the slab detachment benchmark
(Section \ref{sec:slab}).

@@

\section{Mantle melting}\label{sec:melt}
Melt fraction in the mantle depends on temperature ($T$, in °C), pressure ($p$, in \si{\giga\pascal}) and the water content in the melt ($X_{H_2O}$, in wt.\%).
The determination of the melt fraction $M$ is implemented as explained by \citet{Katz2003}, according to the successive modification by \citet{Langmuir2006} and
\citet{Kelley2010}, as follows
\begin{eqnarray}\label{eq:melt_fr}
M(p,T,X_{H_2O})&=&-T+T_{sol}+(x_t\cdot ln(p+y_t))\cdot X_{H_2O}-\nonumber\\
&&+K_m\left(\frac{X_{H_2O}^{bulk}}{D_{H_2O}(1-X_{H_2O})+X_{H_2O}}\right)^{\gamma}
\end{eqnarray}
where $T_{sol}$ is the temperature of the dry solidus, $D_{H_2O}$ is the partition coefficient and $x_t\cdot log(p+y_t)$ indicates the pressure dependence of
the melting curve \citep{Kelley2010}. $x_t$, $y_t$, $K_m$ and $\gamma$ are parameters chosen in according to \citet{Kelley2010}.
However, $X_{H_2O}$ depends on the melt fraction as
\begin{equation}\label{eq:melt_wt}
X_{H_2O}(M)=\frac{X_{H_2O}^{bulk}}{D_{H_2O}+M(1-D_{H_2O})}
\end{equation}
and a numerical solution can be found using a root-finder method \citep{Katz2003,Wang2016}. Furthermore, the content of water in the melt is limited by the
pressure-dependent saturation concentration of water in the melt, determined as
\begin{equation}\label{eq:melt_max}
X_{H_2O}^{sat}=\chi_1 p^{\lambda_m}+\chi_2 p
\end{equation}
where $\chi_1$, $\chi_2$ and $\lambda_m$ are parameters chosen according to \citet{Katz2003}, \citet{Langmuir2006} and \citet{Kelley2010}. The correctness in
the determination of melt fractions and water contents in the melt are tested comparing model results with experimental curves obtained by \citet{Katz2003}
(Sec. \ref{sec:katz}).

@@

\subsection{Instantaneous 2D sphere}\label{sec:ist_sphere}
The domain is a unit square with gravity $\bm{g}=(0,-1)$. The fluid has constant density and viscosity ($\rho_f=1$ and $\eta_f=1$). The sphere is in the middle
of the domain with a radius R = 0.123456798 and has constant density and viscosity ($\rho_s=10^{-2}+\rho_f$ and $\eta_s=10^3 \cdot \eta_f$).
We distinguish three different types of velocity boundary conditions:
\begin{enumerate}
\item FS: free slip conditions on all sides.
\item NS: no slip conditions on all sides.
\item OT: free slip conditions on the sides and the bottom, and open boundary at the top.
\end{enumerate}
All three cases are performed for different grid resolutions, between \num{16x16} and \num{512x512} elements with 50 randomly distributed markers per element,
and different average schemes for the viscosity (harmonic, geometric and arithmetic). The velocity in the centre of the sphere $\bm{v}(0.5;0.5)$), minimum and
maximum velocities ($u_{min}$, $u_{max}$, $v_{min}$, $v_{max}$) and pressures ($p_{min}$, $p_{max}$) on the entire domain, $v_{\textrm{rms}}$ and average pressure on the
entire domain ($p_{avg}$) are compared with solutions generated by ASPECT \citep{KHB12,heister_aspect_methods2,aspect-doi-v2.2.0,aspectmanual} and all results
can be found at \url{https://github.com/cedrict/fieldstone/tree/master/images/stokes_sphere2D} and 
\url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum_equation/Instantaneous_2D_Sphere}. Results in terms of the $v_{\textrm{rms}}$ for NS model 
are also shown in Fig. \ref{fig:inst_sphere}.

@@

\subsection{2D time-dependent Stokes sphere with deformable free surface}\label{sec:sphere}
This experiment is performed in a unit square domain with the gravity acceleration fixed to $g_y=-1$. The sphere has $\rho_s=2$ and $\eta_s=10^3$ and is
initially centred at ($0.5;0.6$) with radius $R=0.123456789$. The fluid surrounding the sphere has $\rho_f=1$ and $\eta_f=1$ and occupies the domain for
$y\leq 0.75$, while for $y>0.75$ the air has $\rho_a=0$ and $\eta_a=10^{-3}$. Velocity boundary conditions are set to free slip on all sides. The Courant
number is set to 0.25. The experiments run for 200 s using grid resolutions from \num{150x150} to \num{512x512} elements, with 25 randomly distributed markers
per element, and for different average schemes for the viscosity. The interface between the fluid and the air is tracked by means of the markers chain. Results
in terms of velocity and pressure fields and topography variations are compared with results obtained with ASPECT
\citep{KHB12,heister_aspect_methods2,aspect-doi-v2.2.0,aspectmanual} and they can be found at
\url{https://github.com/cedrict/fieldstone/tree/master/images/stokes_sphere_fs2D} and
\url{https://github.com/aleregorda/Benchmarks/tree/main/Surface_processes/Time_dependent_sphere}. Results in terms of the $v_{\textrm{rms}}$ in case of 
an arithmetic average are also shown in Fig. \ref{fig:inst_spherefs}.

@@

\subsection{Viscoplastic mantle convection}\label{sec:visco_mantle}
This benchmark is performed as Cases 1-5a described in \citet{Tosi2015} in a 2D unit square domain with two different grid resolutions of \num{32x32} 
and \num{100x100} elements. Reference viscosity, reference density, heat capacity, thermal conductivity and thermal expansion coefficient are fixed to a constant
value of 1 ($\eta_0= \rho_0= C_p = k=\alpha=1$), while the gravity has been chosen as $g_y=10^2$, in order to obtain a Rayleigh number $Ra=10^2$. Temperatures
are set to 0 on top and 1 on bottom of the domain. Velocity boundary conditions are set to free slip on all sides. The initial temperature field is given by
\[T(x,y)=(1-y)+0.01 \cos (\pi x) \sin (\pi y)\].
The viscosity ﬁeld $\eta$ is calculated as the harmonic average between a linear part $\eta_{lin}$ that depends on temperature only or on temperature and 
depth $y$ and a nonlinear-plastic part $\eta_{pl}$ that depends on the strain rate $\dot{\bm{\epsilon}}$, as follows
\[\eta(T,y,\dot{\bm{\epsilon}})=2\left(\frac{1}{\eta_{lin}(T,y)}+\frac{1}{\eta_{pl}(\dot{\bm{\epsilon}})}\right)^{-1}\]
The linear and the nonlinear-plastic parts of the viscosity are calculated following \citet{Tosi2015} as follows
\[\eta_{lin}(T,y)=\exp(-\ln(\Delta\eta_T)T+\ln(\Delta\eta_y)y)\]
\[\eta_{pl}(\dot{\bm{\epsilon}})=\eta^*+\frac{\sigma_Y}{\sqrt{\dot{\bm{\epsilon}}:\dot{\bm{\epsilon}}}}\]
where $\Delta\eta_T$, $\Delta\eta_y$, $\eta^*$ and $\sigma_Y$ are parameters chosen as listed in Table \ref{tab:tosi_setup}.

@@

The experiment is performed with two grid resolution (\num{125x40} and \num{200x80} elements), with 100 markers per element and CFL of 0.25. $v_{\textrm{rms}}$
(see Sec. \ref{sec:error}) calculated for both the simulations match well with results from \citet{vanKeken1997} and obtained with ASPECT
\citep{KHB12,heister_aspect_methods2,aspect-doi-v2.2.0,aspectmanual} and ELEFANT \citep{Thieulot2014}, obtained with similar grid resolutions
(Fig. \ref{fig:thin}). In addition, a variety of simulations with different resolutions and aspect ratios are performed to verify that the sensitivity of the
initial velocity field. Results match well with results from \citet{vanKeken1997} (in the grey area) and \citet{Thieulot2014} (Fig. \ref{fig:thin_initial}). 
All data can be found at \url{https://github.com/aleregorda/Benchmarks/tree/main/Momentum%2BEnergy/Thin_layer}.
